MR-PFU: A suite of R packages for energy conversion chain analysis

### 2023-07-24 paper.Rmd
Editing.

Energy flows through societal energy conversion chains (ECCs)
enable economic activity and facilitate human flourishing.
To understand economic growth and human well-being, 
the field of energy analysis evaluates ECCs
from the primary stage 
(coal, oil, natural gas) 
to the final stage
(energy purchased by consumers, such as refined petroleum and electricity),
to the useful stage
(heat, mechanical drive, and light),
and sometimes to energy services
(transport, thermal comfort, and illumination).
Societal exergy analysis (SEA), an extension of energy analysis,
quantifies ECCs as exergy, 
the mechanical work potential of energy.

@@

Historically, SEA has analyzed the ECCs of individual countries, 
starting with primary and final stage data from the [IEA](https://www.iea.org/data-and-statistics/data-product/world-energy-balances)'s
world energy balances,
using a large number of linked spreadsheets, 
with data in varying and inconsistent formats 
[@Ayres:2003ec; @Serrenho:2014aa; @Brockway:2014aa; @Brockway:2015aa].
The authors of this paper and others in the field
wanted to expand SEA to cover all countries, but
the spreadsheet approach to data analysis was
deemed not scalable.
We created a suite of `R` packages to provide scalable SEA 
for every country in the world.
This paper describes the design of those packages and
demonstrates briefly their use.

@@

Further development followed from selection of the PSUT data format.
First, the PSUT framework (a matrix-based data format) brought the challenge that 
different countries and years 
have varying energy energy carriers (products) and
varying energy conversion machines (industries).
I.e., matrices for different countries and years have differing row and column names.
To get around this challenge, we created an `R` package 
([matsbyname](https://github.com/MatthewHeun/matsbyname/)) which
enables matrix mathematics that respects matrix row and column names, 
inserting `0` rows or columns where needed.
Second,
we knew it would be convenient to perform _matrix_ mathematics 
as easily as  _scalar_ mathematics
in `R` data frames
using the [tidyverse](https://www.tidyverse.org) syntax. 
We developed the [matsindf](https://matthewheun.github.io/matsindf/) package 
to enable this functionality.
Finally, manipulating row and column names proved to be a challenge, 
especially for matrices in PSUT data frames, so 
we developed the [RCLabels](https://matthewheun.github.io/RCLabels/) package
for that purpose.
The table below summarizes these packages, 
all of which are generally useful and available on 
[CRAN](https://cran.r-project.org).

>>>>

Energy flows through societal energy conversion chains (ECCs)
to enable economic activity and facilitate human flourishing.
To understand economic growth and human well-being, 
the field of energy analysis evaluates ECCs
from the primary stage 
(coal, oil, natural gas) 
to the final stage
(energy purchased by consumers, such as refined petroleum and electricity),
to the useful stage
(heat, mechanical drive, and light),
and sometimes to energy services
(transport, thermal comfort, and illumination).
Societal exergy analysis (SEA), an extension of energy analysis,
quantifies ECCs as exergy, 
the mechanical work potential of energy.

@@

Historically, societal exergy analysts have
analyzed the ECCs of individual countries
using a large number of linked spreadsheets,
starting with primary and final stage data from the [IEA](https://www.iea.org/data-and-statistics/data-product/world-energy-balances)'s
world energy balances
[@Ayres:2003ec; @Serrenho:2014aa; @Brockway:2014aa; @Brockway:2015aa]. 
Data were in varying and inconsistent formats.
The authors of this paper and others in the field
wanted to expand SEA to cover all countries, but
the spreadsheet approach to data analysis was
deemed not scalable.
We created a suite of `R` packages to create a database 
of PFU data (the PFU Database)
for all countries in the world from 1960--2020.
The packages enable, for the first time, scalable SEA.
This paper describes the design of those packages and
demonstrates briefly their use.

@@

Further development followed from selection of the PSUT data format.
First, the matrix-based PSUT framework brought the challenge that 
different countries and years 
have varying energy energy carriers (products) and
varying energy conversion machines (industries),
meaning that PSUT matrices for different countries and years 
have differing row and column names.
To get around this challenge, we created the
([matsbyname](https://github.com/MatthewHeun/matsbyname/)) 
package which enables matrix mathematics 
that respects matrix row and column names, 
inserting rows or columns of `0`s when needed.
Second,
we knew it would be convenient to perform _matrix_ mathematics 
as easily as _scalar_ mathematics
in `R` data frames
using the [tidyverse](https://www.tidyverse.org) syntax. 
We developed the [matsindf](https://matthewheun.github.io/matsindf/) package 
to enable this functionality.
Finally, manipulating row and column names proved to be a challenge, 
especially for matrices in PSUT data frames, so 
we developed the [RCLabels](https://matthewheun.github.io/RCLabels/) package
for that purpose.
The table below summarizes these packages, 
all of which are generally useful and available on 
[CRAN](https://cran.r-project.org).
